---
title: "Modeling the Impact of a Household Carbon Tax in Argentina: A Microsimulation Exercise"
author: "Priscila Stitsman, Ines Dezcallar, David Gentili and Asad Azhar"
format: html
editor: visual
---

```{r}
#Library Setup
library(readr)
library(here)
library(tidyverse)
library(ggplot2)
```

## Loading Dataset

```{r}
####  Household Income 
hh_2018 <- read_delim(here("data","engho2018_hogares.txt"),
  delim = "|",
  col_names = FALSE,
  locale = locale(encoding = "Latin1")
)

colnames(hh_2018) <- strsplit(
  "id.provincia.region.subregion.trimestre.anio.pondera.cv1c04.cv1c05_a.cv1c05_b.cv1c05_c.cv1c05_d.cv1c06_e.cv1c06_f.cv1c07.cv1c08.cv1c09.cv1c10.cv1c11.cv1c12.cv1c13.ch01.ch02.ch03.ch04.ch05.ch06.ch07.ch08.ch09.ch10.ch11.ch12.ch13.ch14.ch15.ch16.ch17_a.ch17_b.ch18.ch19.ch20.ch21.propauto.regten.hacina.jniveled.jsexo.jedad_agrup.jsitconyugal.jdif_lp.jestado.jocupengh.jpercept.jcomed.tipohog.cantping.reldep.cantmiem.cantadequi.menor18.menor14.mayor65.dif_lph.cant_act.permieoc.clima_educativo.gastot.gastotpc.gasto_comunh.gascomp.gasvent.gc_01.gc_02.gc_03.gc_04.gc_05.gc_06.gc_07.gc_08.gc_09.gc_10.gc_11.gc_12.gc09_01.gc09_02.gc09_03.gc09_04.gc09_05.gc09_06.gc09_07.gc09_08.gc09_09.fp_contado.fp_credito.fp_especie.fp_prodprop.fp_tarjetas.fp_otras.fp_indef.tn_hipsup.tn_autoserv.tn_negesp.tn_restaurant.tn_kiosco.tn_internet.tn_comedor.tn_especie.tn_otros.tn_indef.decgaphp.decgaphr.decgapht.quigaphp.quigaphr.quigapht.gasto_imputado.m_gasto_imputado.ingtoth.ingpch.m_ingtoth.ingtoth_imp.qinth_p.qinth_r.qinth_t.dinth_p.dinth_r.dinth_t.qinpch_p.qinpch_r.qinpch_t.dinpch_p.dinpch_r.dinpch_t",
  "\\."
)[[1]]

# Count of missing values in each column
#colSums(is.na(hh_2018))

# First rows as colnames
colnames(hh_2018) <- as.character(unlist(hh_2018[1, ]))
hh_2018 <- hh_2018[-1, ]
hh_2018 <- type.convert(hh_2018, as.is = TRUE)

##### Household Expenditure

gastos <- read_delim((here("data",
                           "engho2018_gastos.txt")),
                     delim = "|",
                     locale = locale(encoding = "Latin1"))
##### Codebook

codes <- read_delim(here("data",
  "engho2018_articulos.txt"),
  delim = "|",
  locale = locale(encoding = "Latin1")
)
```

## Cleaning & Wrangling Data

```{r}
########################################################
##### Energy expenditure: electricity, gas, others ##### 
########################################################


energy_exp <- gastos %>%
  filter(articulo %in% c("A0451101", "A0451102", "A0451103",
                         "A0452101", "A0452102", "A0452103", 
                         "A0452104", "A0452105", 
                         "A0453101", "A0453102", "A0453103")) %>%
  mutate(articulo_label = case_when(
    articulo == "A0451101" ~ "Electricidad - vivienda principal",
    articulo == "A0451102" ~ "Electricidad - vivienda secundaria",
    articulo == "A0451103" ~ "Gastos de electricidad realizados para una vivienda ocupada por otro hogar",
    articulo == "A0452101" ~ "Gas envasado en garrafas",
    articulo == "A0452102" ~ "Gas a granel",
    articulo == "A0452103" ~ "Gas natural - vivienda principal",
    articulo == "A0452104" ~ "Gas envasado en tubo",
    articulo == "A0452105" ~ "Gas natural - vivienda secundaria",
    articulo == "A0453101" ~ "Kerosene",
    articulo == "A0453102" ~ "Leña y carbón",
    articulo == "A0453103" ~ "Otros combustibles para el hogar",
    TRUE ~ "Otro"
  ))


###########################
##### Household Income ####
###########################

income_hh <- hh_2018 %>% select(id, ingtoth, ingpch)


## Merging household income with expenditure in energy 

carbon_tax <- energy_exp %>%
  left_join(income_hh %>% select(id, ingtoth), by = "id") %>%
  select (-miembro, -subclase, -clase, -grupo, -division, -forma_pago, -modo_adq, -lugar_adq, -tipo_negocio)

```

## Introduction

#### Motivation

With [394.76 million tonnes](https://www.climatewatchdata.org/countries/ARG?end_year=2021&start_year=1990) of CO₂ emitted in 2021, Argentina accounts for approximately 0.83% of global emissions. Although not a major polluter, the government implemented a tax reform in 2017 that included the introduction of a carbon tax through the amendment of Law No. 23.966. As a result, Argentina is one of the few Latin American countries with a carbon tax, alongside Colombia, Chile, Mexico, and Uruguay. This reform aligns with a growing global trend toward carbon taxation.

Carbon taxes are increasingly seen as a crucial tool to drive environmental sustainability by reducing greenhouse gas emissions. [There are indeed, a cost-effective means of reducing emissions compared to command-and-control regulations or subsidies (Gugler et al., 2021).](https://www.sciencedirect.com/science/article/pii/S0095069620301285) As a Pigouvian tax, carbon pricing helps internalize the external costs of carbon emissions, creating incentives for households to adjust their energy consumption behavior toward cleaner alternatives.

However, literature suggests that they can be regressive, disproportionately burdening lower-income households. This is particularly concerning given the global push toward ecological transition and sustainability. As countries implement carbon tax policies to meet their climate goals, it is essential to assess their potential socioeconomic impacts. Understanding how various income groups will be affected is critical to ensuring that the transition to a sustainable economy is both fair and inclusive. A well-designed policy can help mitigate the risk of exacerbating inequality, ensuring that environmental goals are achieved without disproportionately harming vulnerable populations.

#### Objective

To evaluate the distributional impact and policy design trade-offs of implementing a household-level carbon tax in Argentina, using a microsimulation model based on household energy expenditure data and carbon pricing structures from Argentina and added policies from developed countries.

#### Context

Carbon taxes are difficult to design, as they involve the complex task of putting a price on emissions and valuing different sources of carbon dioxide emissions differently. There are many ways to approach this, which is why carbon taxes are regulated in a variety of ways across countries. For the purposes of our microsimulation, we will consider three different carbon tax models, drawing primarily from two countries. We will take inspiration from the Argentinian model, as well as from Canada, which has implemented a more aggressive carbon tax system.

In 2017, Macri's government introduced a carbon tax, layered on top of an existing tax on liquid fuels. The carbon tax consists of fixed lump sums per unit volume of each fuel, adjusted for inflation every quarter and varying according to the carbon dioxide (CO₂) emissions associated with each fossil fuel. The government planned a gradual increase in the tax rate, an annual[10% hike from 2019 to 2028](https://www.mexico2.com.mx/uploadsmexico/file/Impuesto%20al%20Carbono%20en%20Argentina%20Enero%202023.pdf). As of 2023, the carbon price resulting from these increases corresponded to approximately USD 10 per ton of C[O₂ (USD 10/tCO₂](https://www.mexico2.com.mx/uploadsmexico/file/Impuesto%20al%20Carbono%20en%20Argentina%20Enero%202023.pdf)). Tax rates are relatively low compared to the what is advised by the IMF, which sets the floor for middle-income countries at USD 50/tCO₂ ([source](https://www.cepal.org/en/publications/81239-economic-policy-and-climate-change-carbon-pricing-latin-america-and-caribbean)), but higher than in regional peers such as Chile, Colombia, and Mexico. The tax is applied upstream, targeting fuel importers and producers, rather than directly to households. Its design resembles that of a typical Pigouvian tax.

On the other hand, Canada has a dual carbon pricing system, consisting of a federal fuel charge and a separate system for large industrial emitters. The federal fuel charge is applied to producers, importers, and distributors of fossil fuels, as well as other CO₂-emitting entities such [as air and marine carriers.](https://www.canada.ca/en/revenue-agency/services/tax/excise-taxes-duties-levies/fuel-charge/registration.html)Although the charge is levied upstream, it is often passed on to consumers.

As of [April 2025](https://www.canada.ca/en/environment-climate-change/services/climate-change/pricing-pollution-how-it-will-work/putting-price-on-carbon-pollution.html), the federal fuel charge has been discontinued, but its structure will inform the design of this microsimulation. Like Argentina, Canada sets specific charges [for specific fuel types](https://www.canada.ca/en/department-finance/news/2021/12/fuel-charge-rates-for-listed-provinces-and-territories-for-2023-to-2030.html). However, each province is allowed to implement a higher carbon price than the federal benchmark. Other countries, such as Mexico, also apply differentiated carbon taxes across regions.

Compared to Argentina, Canada enforces a higher and more uniform carbon price (approximately \$50/tCO₂e). Additionally, with the exception of the Yukon and Nunavut territories, Canada provides direct lump-sum rebates to households through the Canada Carbon Rebate scheme. These rebates are adjusted based on family size, rural residence, and province, helping to offset the increase in living costs resulting from the carbon tax. This model offers a valuable benchmark for balancing equity and efficiency in carbon pricing policy design.

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

\*\*Add comment \> argentina, not commited to reduce climate change, need to rise to be effective

------------------------------------------------------------------------

## Microsimulation Model Structure

#### Basic Microsimlation Model

Our basic model will have the following characteristics:

-   Use microdata from Argentina's national household expenditure survey (Encuesta Nacional de Gastos de los Hogares, ENGHo) to estimate baseline household spending on energy sources.

-   Apply carbon price increases to each energy source using the official carbon content conversion table (from Argentina's and Canada's tax authority).

-   Model price pass-through to consumers under the assumption that households ultimately bear the tax burden through higher energy prices.mi

As such we have created three basic model that try to emulate three different carbon taxation systems:

-   A first model that simulates a uniform tax across all households.

-   A second model that includes variation in tax rates depending on how pollutant the energy source used is.

-   A third microsimulation that attempts to mimic the Argentinean carbon tax as it is regulated with lump sums instead of tax rates.

-   Canada?

#### Data Description

In the data frame we have constructed basing ourselves on the 2018 household survey, we have information covering 21,065 households, identified by the variable "id". The data was collected between November of 2017 and November of 2018.

Our data includes spending per household, as well as quantity consumed per household, on the following energy sources:

-   Bottled gas / LPG in canisters (Gas envasado en garrafas)

-   Bulk-delivered LPG (Gas a granel)

-   Gas in larger cylinders or tanks (Gas envasado en tubo)

-   Markatable Natural Gas (Gas Natural)

-   Electricity (Electricidad)

-   Kerosene

-   Other liquid fossil fuels (Otros combustibles)

-   Coal and fuelwood (Leña y Carbón)

#### Sensitivity Analysis

Based on these three different models, a sensitiviy analysis will be carried that will include different scenarios.

##### Elasticity Analysis

Built in the based model: We will conduct a study on how increases in carbon prices affect household demand for different energy sources. To simulate demand elasticity, we will draw on the findings of \*\*\*\*\*\*\*. We will assume that demand elasticities vary across income quintiles, reflecting differing levels of price sensitivity.

##### Different Carbon Price Levels

First, we will include a simulation that includes different price levels, namely a gradual increase in rates to simulate the carbon taxation rates at the levels expected by the IMF

-   Simulate different carbon price levels (e.g., \$10, \$25, \$50/tCO2E) to compare distributional effects under:

    -   A uniform price across all households.

    -   A differential price structure based on income quintiles (e.g., lower rates for lower income households).

##### Distributional Analysis

Furthermore, we will evaluate the impact of carbon taxes as set forth on our three initial model on income distribution. In order to do that will first estimate the post-tax energy expenditures for each household, and then estimate the impact on the tax income and income distribution per income quintiles.

We will evaluate then, in a second \*\*\*, evaluate revenue recycling mechanisms through cash transfers:

-   Flat per capita rebate (uniform transfer to all)
-   Progressive rebate (Enhanced rebates for lower-income households, Higher transfers for rural areas (aligned with Canadian policy))
-   Compare net distributional impacts: tax paid minus rebate received.

##### Aaron's suggestion: Geography Analysis

Will the carbon tax have different effects in different parts of Argentina, as the prevalence of different climates accross the country could influence the impact on household's energy consumption.

## Microsimulation

### Model I: A uniform tax across all households

```{r}
#Suppose we add a 10% tax on the price paid by all households. What would be the final amount paid, the income effect and revenue effect?

# Build function to calculate carbon tax of 10%
calculate_carbon_tax <- function(monto) {
  tax <- monto * 0.10
  return(tax)
}

# Apply function to dataframe
carbon_tax <- carbon_tax %>%
  mutate(
    carbon_tax = map_dbl(
      .x = monto, #apply the function to every value of the column "monto"
      .f = calculate_carbon_tax
    ),
    # Calcular el ingreso neto después del impuesto
    net_income = ingtoth - carbon_tax
  )

# Calculate quintiles
carbon_tax <- carbon_tax %>%
  # Creamos una variable de decil basada en el ingreso total
  mutate(
    quintile = ntile(ingtoth, 5)
  )

# Avg values by quintile
quintile_averages <- carbon_tax %>%
  group_by(quintile) %>%
  summarise(
    avg_total_income = mean(ingtoth, na.rm = TRUE),
    avg_net_income = mean(net_income, na.rm = TRUE),
    avg_tax = mean(carbon_tax, na.rm = TRUE),
    # Calcular el porcentaje que representa el impuesto del ingreso total
    tax_percentage = mean(carbon_tax/ingtoth * 100, na.rm = TRUE)
  )



# Pivot long for plotting
plot_data <- quintile_averages %>%
  select(quintile, avg_total_income, avg_net_income) %>%
  pivot_longer(
    cols = c(avg_total_income, avg_net_income),
    names_to = "income_type",
    values_to = "value"
  )
```

```{r}
### OUTPUTS

# $Income – $energy spend: How much income increases or decreases before and after by quintile.

income_plot_model1 <- ggplot(plot_data, aes(x = factor(quintile), y = value, fill = income_type)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(
    values = c("avg_total_income" = "steelblue", "avg_net_income" = "tomato2"),
    labels = c("Ingreso Total", "Ingreso Neto")
  ) +
  labs(
    title = "Effect of Carbon Tax on Household Income by Quintile",
    subtitle = "Comparison of average total vs. post-tax (net) income across income groups",
    x = "Income Quintile (from lowest to highest)",
    y = "Average Monthly Income (in local currency)",  # adjust unit if needed
    fill = ""
  ) +
  theme_minimal()


# Relative Tax Burden
tax_burden_plot_model1 <- ggplot(quintile_averages, aes(x = factor(quintile), y = tax_percentage)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(
    title = "Carbon Tax as a Share of Income by Quintile",
    subtitle = "Lower-income households bear a larger burden relative to their income",
    x = "Income Quintile (from lowest to highest)",
    y = "Carbon Tax (% of Income)"
  ) +
  theme_minimal()


#Revenue collected by the government
tax_by_quintile <- carbon_tax %>%
  group_by(quintile) %>%
  summarise(
    total_tax = sum(carbon_tax, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    total_revenue = sum(total_tax),
    tax_share = total_tax / total_revenue * 100  # proporción pagada por cada quintil
  )

#Contribution by quintile
tax_share_plot_model1 <- ggplot(tax_by_quintile, aes(x = factor(quintile), y = tax_share)) +
  geom_bar(stat = "identity", fill = "#5A3989") +
  labs(
    title = "Carbon Tax Burden by Income Quintile",
    subtitle = "Share of government revenue from carbon tax contributed by each income group",
    x = "Income Quintile (from lowest to highest)",
    y = "Share of Total Carbon Tax Revenue (%)"
  ) +
  theme_minimal()



print(income_plot_model1)
print(tax_burden_plot_model1)
print(tax_share_plot_model1)
```

### Model II: Tax varies according to energy source, with more pollutant sources with higher taxes, and is consistent with lump sums established by the Argentinian Government.

In this third version of the model, we aim to create a microsimulation that more closely reflects the carbon taxation system currently in place in Argentina.

Since our household survey is from 2018, with data collected in 2017, we will use the lump sums from the first trimester of 2018 as the basis for our calculations. However, because the survey spans from November 2017 to November 2018—a period during which tax rates changed—we will assume that the applicable lump sum is the average of the four quarterly amounts charged throughout 2018.

-   Under Argentina’s carbon tax, only a limited number of fossil fuels are taxed. These include:

-   Unleaded naphtha, up to 92 RON (per litre)

-   Unleaded naphtha, more than 92 RON (per litre)

-   Virgin naphtha (per litre) Petroleum solvent (per litre)

-   White spirit (per litre) Gas oil (per litre)

-   Diesel oil (per litre) Kerosene (per litre)

-   Petroleum coke (per kilogram)

-   Coal (per kilogram)

Among the energy sources used at the household level, only two fall under this taxation scheme: kerosene and the broad category “other liquid fossil fuels.” For these two, we have data on the quantities (in litres) consumed by each household.

Accordingly, I will apply the average of the four quarterly taxes for 2018 to kerosene. For “other liquid fossil fuels,” I will construct an average tax based on the mean of the lump sums applied to all other relevant fuel types. All other energy sources will be considered untaxed under this simulation.

This approach is admittedly imperfect, as it does not account for the share of electricity generated from fossil fuels, notably, coal, which should, in principle, also be subject to carbon taxation. As we don't know the amounts of these energy sources used to generated that a given amount of electricity, we cannot include them in this model.

As a result, a portion of households will remain untaxed in the model, even though their actual energy consumption would indirectly fall under the carbon tax in real life.

```{r}


# Build function to calculate differentiated carbon tax rates based on the energy source

  calculate_carbon_tax <- function(cantidad, articulo) {
 
   lump_sums <- case_when(
     
    #Only Kerosene and Ohter liquid fossil fuels
    
        articulo == "A0453101" ~ mean(0.473, 0.505, 0.549, 0.626),  # Kerosene 
   
        articulo == "A0453103" ~ mean( 0.412, 0.440, 0.478, 0.545, 0.412, 0.440, 0.478, 0.545,0.412, 0.440, 0.478, 0.545,0.412, 0.440, 0.478, 0.545,0.412, 0.440, 0.478, 0.545,0.412, 0.505, 0.549, 0.545, 0.473, 0.505, 0.549, 0.626, 0.473, 0.505, 0.549, 0.626), #Otros Combustibles
        
         TRUE ~ 0
  )
  
  # Calculate tax
  tax <- cantidad * lump_sums
  return(tax)
}

# Step 1: calculate tax by energy type
carbon_tax <- carbon_tax %>%
  mutate(
    carbon_tax_item = map2_dbl(
      .x = cantidad,
      .y = articulo,
      .f = calculate_carbon_tax
    ),
    # Effective tax rate
    tax_rate = carbon_tax_item / monto
  )

# Step 2: Total tax paid by household
household_tax <- carbon_tax %>%
  group_by(id) %>%
  summarise(
    total_carbon_tax = sum(carbon_tax_item, na.rm = TRUE),
    ingtoth = first(ingtoth)  # Take the first income by id (it is the same for each id) 
  ) %>%
  mutate(
    net_income = ingtoth - total_carbon_tax
  )

# Step 3: Calculate quintiles based on income
household_tax <- household_tax %>%
  mutate(
    quintile = ntile(ingtoth, 5)
  )

# Step 4: Analyzing income
quintile_analysis <- household_tax %>%
  group_by(quintile) %>%
  summarise(
    avg_total_income = mean(ingtoth, na.rm = TRUE),
    avg_net_income = mean(net_income, na.rm = TRUE),
    avg_tax = mean(total_carbon_tax, na.rm = TRUE),
    avg_tax_burden = mean(total_carbon_tax/ingtoth * 100, na.rm = TRUE) 
  )

carbon_tax_with_quintile <- carbon_tax %>%
  left_join(
    household_tax %>% select(id, quintile, total_carbon_tax), 
    by = "id"
  )

# Step 5: analyzing quintile tax expenditure 
quintile_fuel_analysis <- carbon_tax_with_quintile %>%
  group_by(quintile, articulo_label) %>%
  summarise(
    avg_carbon_tax = mean(carbon_tax_item, na.rm = TRUE),
    avg_tax_rate = mean(tax_rate * 100, na.rm = TRUE),  
    count = n()  
  )

```

```{r}
#### OUTPUTS
# $Income – $energy spend: How much income increases or decreases before and after by quintile.
income_comparison <- quintile_analysis %>%
  select(quintile, avg_total_income, avg_net_income) %>%
  pivot_longer(
    cols = c(avg_total_income, avg_net_income),
    names_to = "income_type",
    values_to = "amount"
  )

income_plot_model2 <- ggplot(
  income_comparison,
  aes(x = factor(quintile), y = amount, fill = income_type)
) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(
    values = c("avg_total_income" = "steelblue", "avg_net_income" = "tomato2"),
    labels = c("Ingreso Total", "Ingreso Neto")
  ) +
  labs(
    title = "Effect of Carbon Tax on Household Income by Quintile",
    subtitle = "Comparison of average total vs. post-tax (net) income across income groups",
    x = "Income Quintile (from lowest to highest)",
    y = "Average Monthly Income (in local currency)",  # adjust unit if needed
    fill = ""
  ) +
  theme_minimal() 



# Relative Tax Burden
tax_burden_plot_model2 <- ggplot(
  quintile_analysis,
  aes(x = factor(quintile), y = avg_tax_burden)
) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(
    title = "Carbon Tax as a Share of Income by Quintile",
    subtitle = "Lower-income households bear a larger burden relative to their income",
    x = "Income Quintile (from lowest to highest)",
    y = "Carbon Tax (% of Income)"
  ) +
  theme_minimal()



# Gráfico 3: Tasa de impuesto promedio por fuente de energía SHOULD WE DELETE THIS ONE?
# energy_tax_plot <- ggplot(
#   quintile_fuel_analysis %>% 
#     group_by(articulo_label) %>%
#     summarise(avg_rate = mean(avg_tax_rate, na.rm = TRUE)),
#   aes(x = reorder(articulo_label, avg_rate), y = avg_rate, fill = articulo_label)
# ) +
#   geom_bar(stat = "identity") +
#   labs(
#     title = "Tasa promedio de impuesto por tipo de energía",
#     x = "",
#     y = "Tasa de impuesto (%)"
#   ) +
#   coord_flip() +
#   theme_minimal() +
#   theme(legend.position = "none")


# Contribution to gov income revenue

#Revenue collected by the government


total_revenue <- sum(household_tax$total_carbon_tax, na.rm = TRUE)
quintile_contributions <- household_tax %>%
  group_by(quintile) %>%
  summarise(
    total_tax_paid = sum(total_carbon_tax, na.rm = TRUE)
  ) %>%
  mutate(
    percent_contribution = 100 * total_tax_paid / total_revenue
  )

carbon_tax_burden_model2 <- ggplot(quintile_contributions, aes(x = factor(quintile), y = percent_contribution)) +
  geom_bar(stat = "identity", fill = "#5A3989") +
  labs(
 title = "Carbon Tax Burden by Income Quintile",
    subtitle = "Share of government revenue from carbon tax contributed by each income group",
    x = "Income Quintile (from lowest to highest)",
    y = "Share of Total Carbon Tax Revenue (%)"
  ) +
  theme_minimal()


# Show plots
print(income_plot_model2)
print(tax_burden_plot_model2)
print(carbon_tax_burden_model2)






##extend this to be changed: 
 ## the diferences between two > if there is a 300, small difference, focus of the change 
# WHAT IS THIS? - NO idea (ASAD)

```

### Model III: Tax varies according to energy source, with more pollutant sources with higher taxes, following the Canadian Model

#### Description

In this second model, we aim to recreate a more plausible carbon tax system, drawing inspiration from Canada’s approach. Like Argentina, Canada sets absolute rates in Canadian dollars per specific volume or weight of various fossil fuels, with more polluting fuels taxed at higher rates than less polluting ones. It establishes specific rates for the following fossil fuels ([source](https://www.canada.ca/en/revenue-agency/services/forms-publications/publications/fcrates/fuel-charge-rates.html#fcrts)):

-   Aviation gasoline

-   Aviation turbo fuel

-   Butane

-   Coke

-   Coke oven gas

-   Combustible waste

-   Ethane

-   Gas liquids

-   Gasoline

-   Heavy fuel oil

-   Kerosene

-   High heat value coal

-   Kerosene

-   Light fuel oil

-   Low heat value coal

-   Marketable natural gas

-   Methanol

-   Naphtha

-   Non-marketable natural gas

-   Petroleum coke

-   Pentanes plus

-   Propane

-   Still gas.

As in Argentina, fuel charge rates in Canada have been gradually increasing. The first carbon price introduced in Canada was set at CAD 20/CO₂t in 2019. Rates increased by CAD 10/CO₂t annually until 2022, and by CAD 15/CO₂t each year from 2023 to 2030. Thus, in 2024, the rate reached CAD 75/CO₂t, which is approximately USD 55.55/CO₂t. This is in line with the level at which middle-income countries should be pricing carbon, according to the IMF (USD 50/CO₂t) ([source](https://www.canada.ca/en/environment-climate-change/services/climate-change/pricing-pollution-how-it-will-work/carbon-pollution-pricing-federal-benchmark-information.html)).

For the purposes of this study, we make Canada’s fuel charges comparable to those of Argentina. Using the inflation calculator provided by the Bank of Canada ([source](https://www.bankofcanada.ca/rates/related/inflation-calculator/)), we calculate a price increase of 20.24% between 2018 and 2024. Accordingly, we adjust the 2024 fuel charge rates to their 2018 equivalents. We then apply the average 2018 exchange rate between the Argentine peso and the Canadian dollar (1 Argentine Peso = 0.04856 Canadian Dollar ([source](https://www.exchange-rates.org/exchange-rate-history/ars-cad-2018))) to convert the rates into Argentine pesos. It should noted that the exchange rate signficantly fluctuate during this year.

Since not all of the energy sources mentioned above are included in this study, we will group the fossil fuels into broader categories based on the energy sources covered in this survey:

-   Bottled gas / LPG in canisters (Gas envasado en garrafas): Propane, Butane, Pentanes plus

-   Bulk-delivered LPG (Gas a granel):Propane, Butane Pentanes plus

-   Gas in larger cylinders or tanks (Gas envasado en tubo)*:* Propane, Butane, Pentanes plus

-   Markatable Natural Gas (Gas natural): Natural Gas

-   Kerosene

-   Other liquid fossil fuels: Gasoline, Heavy fuel oil, Light fuel oil, Methanol. Naphta

-   Coal: High heat value coal, Low heat value coal, Coke, Coke oven gas. Since these fuels are rarely used by households bur rather in industrial processes, and the charges associated with them are abnormally high, they will not be included in this study.

As we did in the case of Argentina, we will calculate an average rate per category, corresponding to the mean of the fuel charge rates applied to the relevant energy source usually used by households. For example, the tax applicable to bottled gas will be equal to the average of the rates for Propane, Butane, and Pentanes Plus. Once we have the average rates, we will apply the conversions described above to obtain values that can be compared to Argentina’s 2018 rates.

#### Code

```{r}
#Creating fuel charge rates for the different categories of energy sources including in the data by averaging Canada's 2024-2025 fuel charges rates

Canada_rates <- data.frame(
  energy_source= c("Propane", "Butane", "Pentanes plus", "Natural Gas", "Kerosene","Gasoline", "Heavy fuel oil", "Light fuel oil", "Methanol", "Naphta"), 
  rates= c(0.1238, 0.1424, 0.1424, 0.1525, 0.2065, 0.1761, 0.2550, 0.2139, 0.0878, 0.1803))%>%
  mutate(
    rates = rates/1.2024
  )%>%
  mutate(
    rates = rates/0.04856
  )

Canada_rates


#Averaging these rates into the different energy sources included in the survey
Canada_rate_per_Energy_source_in_survey <- Canada_rates%>%
  filter(energy_source == "Propane")%>%
  mutate (
    Gas_1_Garrafas = mean(c(2.120279, 2.438835, 2.438835 )), 
    Gas_2_Granel = mean(c(2.120279, 2.438835, 2.438835 )),
    Gas_3_Tubo = mean(c(2.120279, 2.438835, 2.438835 )),
    Gas_4_Natural = 2.611814, 
    Kerosene = 3.536653,
    Liquid_Fuels = mean(c(3.016003, 4.367295, 3.663390, 1.503720, 3.087935))
  )%>%
  select(-rates, -energy_source)%>%
  pivot_longer(cols = everything(), 
               names_to = "energy_type", 
               values_to = "rate")

elasticity <- 0   # Change for different scenarios
rebate_mode <- "none"  # Options: "none", "uniform", "by_quintile"
rebate_target_quintiles <- c(1, 2)  # Only used if rebate_mode == "by_quintile"
gas_increment <- 0 # Change ratio to increase / decrease tax on gas articles
fossil_increment <- 0 # Change ratio to increase / decrease tax on fossil articles
rebate_share <- 1  # Ratio of revenue that goes into rebate (can't be more than 1)


# Function that applies differentiated tax rates by type of item/energy
calculate_carbon_tax <- function(cantidad, articulo, 
                                 fossil = fossil_increment, #ratio of increment 
                                 gas = gas_increment) #ratio of increment 
{
  # Tax rates based on type of energy
  tax_rates <- case_when(
    # Highly polluting fossil fuels
    articulo == "A0453101" ~ 3.54* (1 + fossil),  # Kerosene 
    articulo == "A0453103" ~ 3.02* (1 + fossil),  # Other fuels 
    
    # Gas (medium emissions)
    articulo == "A0452101" ~ 2.12 * (1 + gas),  # Bottled gas in cylinders 
    articulo == "A0452102" ~ 2.12 * (1 + gas),  # Bulk gas
    articulo == "A0452103" ~ 2.61 * (1 + gas),  # Natural gas - primary residence
    articulo == "A0452104" ~ 2.12 * (1 + gas),  # Gas in tubes
    articulo == "A0452105" ~ 2.61 * (1 + gas),  # Natural gas - secondary residence 
    
    # Electricity (varies by energy matrix, but directly less polluting)
    articulo == "A0451101" ~ 0,  # Electricity - primary residence
    articulo == "A0451102" ~ 0,  # Electricity - secondary residence
    articulo == "A0451103" ~ 0,  # Electricity expenses for another residence
    
    # Firewood and coal (may be renewable or not depending on the source)
    articulo == "A0453102" ~ 0,  # Firewood and coal 
    
    # Others
    TRUE ~ 0                     # Standard rate 
  )
  
  # Calculate tax
  tax <- cantidad * tax_rates
  return(tax)
}


# Step 1: calculate tax by energy type
carbon_tax <- carbon_tax %>%
  mutate(
    carbon_tax_item = map2_dbl(
      .x = monto,
      .y = articulo,
      .f = calculate_carbon_tax 
    ),
    # Effective tax rate
    tax_rate = carbon_tax_item / monto,
    
    # Apply elasticity adjustment to quantity 
    monto = monto * (1 + tax_rate) ^ elasticity,
    carbon_tax_item = monto * tax_rate  
  )

# Step 2: Total tax paid by household
household_tax <- carbon_tax %>%
  group_by(id) %>%
  summarise(
    total_carbon_tax = sum(carbon_tax_item, na.rm = TRUE),
    ingtoth = first(ingtoth), # Take the first income by id (it is the same for each id) 
    total_spent_energy = sum(monto, na.rm = TRUE) 
  ) %>%
  mutate(
    net_income = ingtoth - total_carbon_tax
  )

# Step 3: Calculate quintiles based on income
household_tax <- household_tax %>%
  mutate(
    quintile= ntile(ingtoth, 5)
  )

## Step 3.1: Add rebate if any

total_revenue <- sum(household_tax$total_carbon_tax, na.rm = TRUE)
rebate_pool <- total_revenue * rebate_share

household_tax <- household_tax %>%
  mutate(
    rebate = case_when(
      rebate_mode == "uniform" ~ rebate_pool / n(),
      rebate_mode == "by_quintile" & quintile %in% rebate_target_quintiles ~
        rebate_pool / sum(quintile %in% rebate_target_quintiles),
      rebate_mode %in% c("by_quintile", "none") ~ 0,
      TRUE ~ 0
    ),
    net_income = net_income + rebate  # updated post-rebate income
  )

# Step 4: Analyzing income
quintile_analysis <- household_tax %>%
  group_by(quintile) %>%
  summarise(
    avg_total_income = mean(ingtoth, na.rm = TRUE),
    avg_net_income = mean(net_income, na.rm = TRUE),
    avg_tax = mean(total_carbon_tax, na.rm = TRUE),
    avg_tax_burden = mean(total_carbon_tax/ingtoth * 100, na.rm = TRUE) 
  )

carbon_tax_with_quintile <- carbon_tax %>%
  left_join(
    household_tax %>% select(id, quintile, total_carbon_tax), 
    by = "id"
  )

# Step 5: analyzing quintile tax expenditure 
quintile_fuel_analysis <- carbon_tax_with_quintile %>%
  group_by(quintile, articulo_label) %>%
  summarise(
    avg_carbon_tax = mean(carbon_tax_item, na.rm = TRUE),
    avg_tax_rate = mean(tax_rate * 100, na.rm = TRUE),  
    count = n()  
  )

```

#### Outputs
```{r}
#### OUTPUTS
# $Income – $energy spend: How much income increases or decreases before and after by quintile.
income_comparison <- quintile_analysis %>%
  select(quintile, avg_total_income, avg_net_income) %>%
  pivot_longer(
    cols = c(avg_total_income, avg_net_income),
    names_to = "income_type",
    values_to = "amount"
  )

income_plot_model2 <- ggplot(
  income_comparison,
  aes(x = factor(quintile), y = amount, fill = income_type)
) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(
    values = c("avg_total_income" = "steelblue", "avg_net_income" = "tomato2"),
    labels = c("Ingreso Total", "Ingreso Neto")
  ) +
  labs(
    title = "Effect of Carbon Tax on Household Income by Quintile",
    subtitle = "Comparison of average total vs. post-tax (net) income across income groups",
    x = "Income Quintile (from lowest to highest)",
    y = "Average Monthly Income (in local currency)",  # adjust unit if needed
    fill = ""
  ) +
  theme_minimal() 



# Relative Tax Burden
tax_burden_plot_model2 <- ggplot(
  quintile_analysis,
  aes(x = factor(quintile), y = avg_tax_burden)
) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(
    title = "Carbon Tax as a Share of Income by Quintile",
    subtitle = "Lower-income households bear a larger burden relative to their income",
    x = "Income Quintile (from lowest to highest)",
    y = "Carbon Tax (% of Income)"
  ) +
  theme_minimal()



# Gráfico 3: Tasa de impuesto promedio por fuente de energía SHOULD WE DELETE THIS ONE?
# energy_tax_plot <- ggplot(
#   quintile_fuel_analysis %>% 
#     group_by(articulo_label) %>%
#     summarise(avg_rate = mean(avg_tax_rate, na.rm = TRUE)),
#   aes(x = reorder(articulo_label, avg_rate), y = avg_rate, fill = articulo_label)
# ) +
#   geom_bar(stat = "identity") +
#   labs(
#     title = "Tasa promedio de impuesto por tipo de energía",
#     x = "",
#     y = "Tasa de impuesto (%)"
#   ) +
#   coord_flip() +
#   theme_minimal() +
#   theme(legend.position = "none")


# Contribution to gov income revenue

#Revenue collected by the government


quintile_contributions <- household_tax %>%
  group_by(quintile) %>%
  summarise(
    total_tax_paid = sum(total_carbon_tax, na.rm = TRUE)
  ) %>%
  mutate(
    percent_contribution = 100 * total_tax_paid / total_revenue
  )

carbon_tax_burden_model2 <- ggplot(quintile_contributions, aes(x = factor(quintile), y = percent_contribution)) +
  geom_bar(stat = "identity", fill = "#5A3989") +
  labs(
 title = "Carbon Tax Burden by Income Quintile",
    subtitle = "Share of government revenue from carbon tax contributed by each income group",
    x = "Income Quintile (from lowest to highest)",
    y = "Share of Total Carbon Tax Revenue (%)"
  ) +
  theme_minimal()


# Show plots
print(income_plot_model2)
print(tax_burden_plot_model2)
print(carbon_tax_burden_model2)

```

